---
title: "An overview of dplyr"
subtitle: "Daryn Ramsden"
author: "thisisdaryn at gmail dot com"
date: "last updated: `r Sys.Date()`"
institution: ""
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: libs/switch-themer.css
    chakra: libs/remark.js
    nature:
      highlightLines: true
      countIncrementalSlides: false
    includes:
      after_body: libs/toggle.html
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      comment = "")
xaringanExtra::use_tile_view()
#xaringanExtra::use_panelset()
xaringanExtra::use_webcam()
#xaringanExtra::use_editable()
xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         
  mute_unhighlighted_code = TRUE  
)

#after installing these packages, please comment ount. 
#install.packages("palmerpenguins")
#install.packages("kableExtra")
#install.packages("xaringan")
#install.packages("devtools")
#devtools::install_github("gadenbuie/xaringanExtra")

library(xaringan)
library(xaringanExtra)
library(palmerpenguins)
library(dplyr)
library(kableExtra)
```



### The data we will be using 


```{r}
#install.packages("palmerpenguins")
library(palmerpenguins)
```

```{r echo = FALSE, eval = FALSE, fig.cap= "Palmer penguins data table."}
DT::datatable(penguins, 
         extensions = c('FixedColumns',"FixedHeader"),
          options = list(scrollX = TRUE, 
                         paging=TRUE,
                         fixedHeader=TRUE,
                         pageLength = 15))
```

```{r echo = FALSE}
library(rmarkdown)
paged_table(penguins)
```




---
### What do these variables represent?

Data were collected and made available by Dr. Kristen Gorman and the Palmer Station, Antarctica LTER, a member of the Long Term Ecological Research Network.

  * *species*: *Adelie*, *Chinstrap* or *Gentoo*
  
  * *island*: *Biscoe*, *Dream* or *Torgersen* (factor)
  
  * *bill_length_mm*: bill length mm (numeric) 
  
  * *bill_depth_mm*: bill depth in mm (numeric)      
  
  * *flipper_length_mm*: flipper length in mm (numeric)
  
  * *body_mass_g*: body mass in grams (numeric)
  
  * *sex*: *male* or *female* (factor)        
  
  * *year*: 2007, 2008 or 2009
  
  
---
## dplyr: a package for data manipulation

The data you get is almost in the form you want

--

`dplyr` is an R package that encapsulates many common data manipulation tasks

--

Sometimes you want to: 
--

  * keep only some of the rows

--

  * keep only some of the columns

--

  * adds new columns

--

  * sort data
  
--

  * provide summary statistics

--

`dplyr` has functions for each of these (and many others)


---
## Using `dplyr`

#### How do you install `dplyr`?

```{r eval = FALSE, fig.cap = "A call to the install.packages function illustrating how to install the dplyr package from CRAN"}
install.packages("dplyr")
# or install.packages("tidyverse)
```


#### How do you use `dplyr`?

```{r fig.cap = "A call to the library function to load the dplyr package"}
library(dplyr)
# or library(tidyverse)
```


---
## Key single table verbs/functions


* Working with rows:

  * `filter`: keep only some of the rows based on column values
  
  * `slice`: keep some of the rows based on their location
  
  * `arrange`: sort data


* Working with columns:

  * `select`: keep only some of the columns 

  * `mutate` adds new columns
  
  * `rename` change the name of specified columns

  * `relocate` changes the order of the columns


* Groups of rows:


  * `summarise` (and `group_by`): provide summary statistics 

  
  
---
## `filter`

#### a function for specifying which rows to keep


Example 1: How do we get all penguins of the Chinstrap species?



---
## `filter`

#### a function for specifying which rows to keep


Example 1: How do we get all penguins of the Chinstrap species?

```{r eval = FALSE, message = FALSE, fig.cap= "A tibble showing the data filtered by chinstrap."}
chinstrap <- filter(penguins, species == "Chinstrap")
```


---
## `filter`

#### a function for specifying which rows to keep


Example 1: How do we get all penguins of the Chinstrap species?

```{r message = FALSE}
chinstrap <- filter(penguins, species == "Chinstrap")
chinstrap
```


---
## `filter`

#### a function for specifying which rows to keep


Example 2: How do we get penguins that are 4 kg or greater?

  
---
## `filter`

#### a function for specifying which rows to keep


Example 2: How do we get penguins that are 4 kg or greater?


```{r eval = FALSE}
penguins_4k <- filter(penguins, body_mass_g >= 4000)
```
---
## `filter`

#### a function for specifying which rows to keep


Example 2: How do we get penguins that are 4 kg or greater?


```{r, fig.cap= "A tibble applying a filter in which penguins are 4kg or greater."}
penguins_4k <- filter(penguins, body_mass_g >= 4000)
penguins_4k
```


---
### Assessment

How many penguins were found on Torgersen island (<i>Torgersen</i>)?


  
---
### Assessment

How many penguins were found on Torgersen island (<i>Torgersen</i>)?



```{r}
torgersen<- filter(penguins, island == "Torgersen")
dim(torgersen)
```

--

Also could have used:


```{r}
torgersen<- penguins %>% filter(island == "Torgersen")
dim(torgersen)
```




---
## `select`

#### A function/verb for specifying which columns to keep 

As of dplyr 1.0 there are 5 ways to use select 

  1. By **position** 
  
  2. By **name**
  
  3. by **function of name** 
  
  4. by **type** 
  
  5. by combination of the above using logical operators (`|`, `&`, `!`)



---
### `select` by position

Example: select columns 1, 3 and 5 from `penguins`

--
```{r eval = FALSE}
penguins %>% select(1, 3, 5)
```


---
### `select` by position

Example: select columns 1, 3 and 5 from `penguins`

```{r, fig.cap= "A tibble in which columns one, three, and five are selected."}
penguins %>% select(1, 3, 5)
```


---
### `select` by name

Example: select *species*, *island* and *body_mass_g*

--
```{r eval = FALSE}
penguins %>% select(species, island, body_mass_g)
```

---
### `select` by name


Example: select *species*, *island* and *body_mass_g*

```{r fig.cap= "A tibble in which species are selected by the species, island, and body_mass_g variables."}
penguins %>% select(species, island, body_mass_g)
```

---
### `select` by a function of column names

`select` can be used in conjunction with other useful functions such as:

  
  * `starts_with` 
  
  * `ends_with`
  
  * `contains` 
  
  * `matches`


---
### `select` by a function of column names


Example: Choose all columns that contain "mm":

```{r}
penguins_mm <- penguins %>% select(contains("mm"))
```


---
### `select` by a function of column names


Example: Choose all columns that contain "mm":

```{r fig.cap= "A tibble in which only includes data that contains mm in the variable names."}
penguins_mm <- penguins %>% select(contains("mm"))

penguins_mm
```


---
### `select` by a function of column names


Example: How to choose all columns starting with "bill":

```{r}
bills_df <- penguins %>% select(starts_with("bill"))
```

---
### `select` by a function of column names


Example: How to choose all columns starting with "bill":

```{r fig.cap= "A tibble that contains only the columns starting with the word bill."}
bills_df <- penguins %>% select(starts_with("bill"))

bills_df
```


---
### `select` by type 

Example: choose all numeric columns:

```{r eval = FALSE}
penguins %>% select(where(is.numeric))
```

---
### `select` by type 

Example: choose all numeric columns:

```{r fig.cap = "A tibble in which all numeric columns are selected."}
penguins %>% select(where(is.numeric))
```


---
### `select` by logical combination

Example: choose all factor variables or variables containing the word "bill"

```{r eval = FALSE}
penguins %>% select(where(is.factor) | contains("bill"))
```



---
### `select` by logical combination

Example: choose all factor variables or variables containing the word "bill"

```{r fig.cap= "A tibble in which all factor variables are seleted or contains the word bill."}
penguins %>% select(where(is.factor) | contains("bill"))
```




---
## `mutate` 

#### a function to add new columns


Example: Adding a column that indicates whether a penguin has a mass greater than 4 kg

```{r eval = FALSE}
penguin_extra <- penguins %>% 
  mutate(above_4kg= if_else(body_mass_g > 4000, TRUE, FALSE))
```


---
## `mutate` 

#### a function to add new columns


Example: Adding a column that indicates whether a penguin has a mass greater than 4 kg

```{r fig.cap = "A tibble which adds a new column called above_kg, which indicates whether a penguin has a mass greater than 4kg."}
penguin_extra <- penguins %>% 
  mutate(above_4kg= if_else(body_mass_g > 4000, TRUE, FALSE))

head(penguin_extra)
```


---
## `arrange` 

#### A function for sorting data

Example: Sort all penguins by body mass:

--

```{r}
penguins_sorted <- penguins %>% arrange(body_mass_g)
```

---
## `arrange` 

#### A function for sorting data

Example: Sort all penguins by body mass:



```{r eval = FALSE,fig.cap= "A data table which shows penguins sorted by ascending body mass."}
penguins_sorted <- penguins %>% 
  arrange(body_mass_g)
penguins_sorted
```
```{r echo = FALSE}
paged_table(penguins_sorted)
```



---
### sorting with multiple columns using `arrange`

Example sorting by species, then by descending order of mass:

```{r eval = FALSE}
penguins_sorted2 <- penguins %>% 
  arrange(species, desc(body_mass_g))
penguins_sorted2
```
```{r echo = FALSE, fig.cap = "A data table which shows penguins sorted by descending body mass."}
penguins_sorted2 <- penguins %>% 
  arrange(species, desc(body_mass_g))
paged_table(penguins_sorted2)
```





---
## `summarise`/`summarize`


#### A verb/function to get summary statistics.


Question: what's the mean flipper length and body mass among the Palmer penguins?


```{r fig.cap= "A tibble that shows the mean flipper length and body mass among the Palmer penguins."}
penguins %>% 
  summarise(num_penguins = n(),
            avg_mass = mean(body_mass_g, na.rm = TRUE),
            avg_fl_length = mean(flipper_length_mm, na.rm = TRUE))
```


---
## `group_by` 

#### A function that makes `summarise` really powerful 

`group_by` creates a grouped data frame based on columns you specify

--

For example, grouping the penguins by island and species:

--

```{r}
gr_penguins <- penguins %>% group_by(island, species)
```


---
## `group_by` 

#### A function that makes `summarise` really powerful 

`group_by` creates a grouped data frame based on columns you specify



For example, grouping the penguins by island and species:

```{r fig.cap = "A tibble that groups the penguins by island and species."}
gr_penguins <- penguins %>% group_by(island, species)
head(gr_penguins)
```





---
## How is the grouped data frame different?

--

  * Extra information is added to the data frame 
  
--


  * rows that match on all the grouping variables will be in the same group

--
  
  * rows that don't match on all the grouping variables will be in different groups
  


---
## `group_by` and `summarise` together

Now let's do the same summary as before with the grouped data:

--

```{r fig.cap= "A tibble that shows the number of penguins by average mass, average flipper length, island, and species."}
gr_penguins %>% summarise(num_penguins = n(),
                       avg_mass = mean(body_mass_g, na.rm = TRUE),
                       avg_fl_length = mean(flipper_length_mm, 
                                            na.rm = TRUE))
```


---
### New features of `summarise`

`dplyr` 1.0 has some new features of `summarise`:

  * summaries that return multiple values
  
  * summaries that return multiple columns
  
  
  
---
### Summaries with multiple values

Example: using `summarise` to get the range of bill lengths for each species of penguin:

```{r fig.cap = "A tibble with 2 columns. The 1st column has each species name in two consecutive rows. The 2nd column features the minimum and maximum bill lengths for each species in alternating rows."}
penguins %>% 
  group_by(species) %>%
  summarise(rng = range(bill_length_mm, na.rm = TRUE))

```

---
### Summaries with multiple columns

Example: using `summarise` to find the minimum and maximum mass penguin on each island:

```{r fig.cap = "A tibble with 3 columns. The first column contains the names of the islands. The 2nd column contains the minimum mass of a penguin on the corresponding island. The 3rd column contains the maximum mass of a penguin on the island." }
penguins %>% 
  group_by(island) %>%
  summarise(tibble(min_mass = min(body_mass_g, na.rm = TRUE),
                   max_mass = max(body_mass_g, na.rm = TRUE)))
```

  



---
### So ... a couple other things about groups

  * default behavior is to remove the last level of grouping after a call to `summarise` 
  
  * grouped data can be used with other `dplyr` verbs e.g. `mutate`
  
  * you can ungroup data using `ungroup`
  
---
### Example using `group_by` with `mutate`

What if we wanted to give each penguin a number within its species? 

```{r}
numbered_penguins <- penguins %>% 
  group_by(species) %>%
  mutate(penguin_num = 1:n())
```

---
### Example using `group_by` with `mutate`

What if we wanted to give each penguin a number within its species? 

```{r fig.cap = "A tibble that adds the field penguin_num, which adds a number to a penguin within its species. In addition, the tibble contains information about the species, island, bill length, bill depth, flipper length, and body mass." }
numbered_penguins <- penguins %>% 
  group_by(species) %>%
  mutate(penguin_num = 1:n())

numbered_penguins
```



---
## `rename`

#### A function/verb to rename columns

Works like `select`


Example: renaming by position

```{r fig.cap = "A tibble that renames column three and four to bill_length and bill_depth. In addition, the tibble shows species, island, flipper length, body mass, and sex."}
penguins_different <- penguins %>% rename(bill_length = 3, 
                                          bill_depth = 4)

penguins_different
```



---
### `rename_with`

`rename_with` can be used with a specified transformation (and optionally with a column selection). 


Example: rename all columns to be uppercase

```{r fig.cap = "A tibble that renames the columns, species, island, bill length, bill depth, flipper length, and body mass to uppercase characters."}
penguins %>% rename_with(toupper)
```



---
## `rename_with`
```{r fig.cap= "This tibble shows column names that are only capitalized to uppercase when the columns contain numeric values."}
penguins %>% rename_with(toupper, where(is.numeric))
```



---
## `relocate`

### A function 


  * (**default**) move selected variables to the front
  
  * move selected columns before a specified location
  
  * move selected columns after a specified location



---
## `relocate` examples

Example: bring all the factor variables to the front 

```{r eval = FALSE}
penguins %>% relocate(where(is.factor))
```


---
## `relocate` examples

Example: bring all the factor variables to the front 

```{r fig.cap = "This tibble brings all the factor variables, which are species, island, and sex, to the front of the tibble."}
penguins %>% relocate(where(is.factor))
```



---
## `relocate` examples

Example: relocate all factor variables after *body_mass_g*  


```{r fig.cap= "This tibble relocates all factor variables to the front after the  body_mass_g column"}
penguins %>% relocate(contains("bill"), .after = body_mass_g)
```


---
### `across`: a really useful new function

What if you wanted the average value - per group - of each numeric column? 



Annoying way: 

```{r eval = FALSE}
penguins %>% group_by(species) %>% 
  summarise(avg_bill_length = mean(bill_length_mm, na.rm = TRUE),
            avg_bill_depth = mean(bill_depth_mm, na.rm = TRUE), 
            avg_fl_length_mm = mean(flipper_length_mm, na.rm = TRUE),
            avg_body_mass_g = mean(body_mass_g, na.rm = TRUE))
```


---
### `across`: a really useful new function

What if you wanted the average value - per group - of each numeric column? 



Annoying way: 

```{r fig.cap = "This tibble shows the average values of bill length, bill depth, flipper length, and body mass."}
penguins %>% group_by(species) %>% 
  summarise(avg_bill_length = mean(bill_length_mm, na.rm = TRUE),
            avg_bill_depth = mean(bill_depth_mm, na.rm = TRUE), 
            avg_fl_length_mm = mean(flipper_length_mm, na.rm = TRUE),
            avg_body_mass_g = mean(body_mass_g, na.rm = TRUE))
```


---
### `across`: a really useful new function

What if you wanted the average value - per group - of each numeric column? 


Neater/better way: 

```{r eval = FALSE}
penguins %>% group_by(species) %>%
  summarise(across(where(is.numeric) & !contains("year"),
                   mean, na.rm = TRUE))
```

---
### `across`: a really useful new function

What if you wanted the average value - per group - of each numeric column? 


Neater/better way: 

```{r fig.cap = "A tibble which uses the across function to calculate the average values of bill length, bill depth, flipper length, and body mass."}
penguins %>% group_by(species) %>%
  summarise(across(where(is.numeric) & !contains("year"),
                   mean, na.rm = TRUE))
```


---
### `across`: a closer look


`across` has two primary arguments:

  * <tt>.cols</tt> selects the columns you want to operate on
  
  * <tt>.fns</tt> is a function or list of functions that you want to apply
     
     * can be a `purrr` style formula 
  



---
### multiple summaries with `across`

Example: For each island, what is the average of all numeric variables and the count of all factor variables?

```{r fig.cap= "A tibble which shows the mean of all the values along with the count of all the factor variables."}
penguins %>%
  group_by(island) %>% 
  summarise(
    across(where(is.numeric), mean, na.rm = TRUE), 
    across(where(is.factor), n_distinct),
    n = n(), 
  )
```


---
### `across` example with `filter`

Example: get all rows without missing values:


```{r}
penguins_complete <- penguins %>% 
  filter(across(everything(), ~ !is.na(.x)))
```

--

Is that any different to?

```{r eval = FALSE}
penguins_complete2 <- penguins %>% 
  filter(across(everything(), complete.cases))
```


---
### `across` example with `distinct`

All combinations of variables meeting specified criteria using `distinct`

```{r fig.cap = "A tibble that shows the distinct combinations across variables which are factors which are species, island, and sex."}
penguins %>% distinct(across(is.factor, sort = TRUE))
```

---
### `across` example with `count`

Counts of all combinations of variables meeting specified criteria using `count`

```{r fig.cap = "A tibble that gives the counts of all combinations of variables that are factors." }
penguins %>% count(across(is.factor, sort = TRUE))
```

---
### `across` example with `mutate`

Using `across` with `mutate` to rescale all numeric variables between 0 and 1


```{r fig.cap= "A tibble which rescale all the numeric variables between 0 and 1."}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

penguins_rescaled <- penguins %>% 
  mutate(across(where(is.numeric), rescale01))

penguins_rescaled
```


---
## Row-wise operations

Question: what if we wanted to create a new column that was the average of the <i>bill_depth_mm</i> and <i>bill_length_mm</i> variables?

You might try:
```{r eval = FALSE}
penguins %>% select(contains("bill")) %>% 
  mutate(avg = mean(c(bill_length_mm, bill_depth_mm), na.rm = TRUE))
```
---
## Row-wise operations

Question: what if we wanted to create a new column that was the average of the <i>bill_depth_mm</i> and <i>bill_length_mm</i> variables?

You might try:
```{r fig.cap="a tibble showing a new column of calculated means but all the values are the same in the new variable"}
penguins %>% select(contains("bill")) %>% 
  mutate(avg = mean(c(bill_length_mm, bill_depth_mm), na.rm = TRUE))
```


---
### Using `rowwise`

We can use `rowwise` prior to mutate instead 

```{r eval = FALSE}
penguins %>% 
  select(contains("bill")) %>%
  rowwise() %>% 
  mutate(avg = mean(c(bill_length_mm, bill_depth_mm), na.rm = TRUE))
```


---
### Using `rowwise`

We can use `rowwise` prior to mutate instead 

```{r fig.cap = "A tibble showing a new column correctly displaying the mean for each row within the row"}
penguins %>% 
  select(contains("bill")) %>%
  rowwise() %>% 
  mutate(avg = mean(c(bill_length_mm, bill_depth_mm), na.rm = TRUE))
```


---
## Joins


To illustrate the join functions, we will use two small data sets

First, a data frame containing the populations of 8 countries (via census.gov): 

```{r fig.cap = "A tibble showing the names and populations of 8 countries"}
populations <- readr::read_csv("data/populations.csv")
populations
```

---
## Joins

Next, a data frame containing the land areas of some countries (via wikipedia)


```{r fig.cap = "A tibble showing the names and land areas of 7 countries"}
areas <- readr::read_csv("data/areas.csv")
areas
```
Note that some countries are in both data frames while others are only in one.

---
### Inner joins with `inner_join`

Inner joins combine tables, taking only entries that are in both:

```{r fig.cap = "A tibble showing the names, areas and populations of 5 countries"}
inner_join(populations, areas)
```



---
### Full joins with `full_join`

Full joins combine tables, taking all entries from either:

```{r fig.cap = "A tibble showing the names, populations and areas of 10 countries with some values missing"}
full_join(populations, areas)
```

---
### Left (or right) joins with `left_join` (or `right_join`)

Left joins take all the rows in the first table along with any rows in the second table that match

```{r fig.cap = "A tibble showing the names, populations and areas of 8 countries. 3 values are missing in the area column."}
left_join(populations, areas)
```